
# Tumor-Stroma Interactions in Multiple Myeloma {#Myeloma}

Cancer cells and stromal cells cooperate by exchanging diffusible factors that 
sustain tumor growth. In the case of multiple myeloma, there are three types
of cells: Malignant plasma cells (MM), osteoblast (OB) and osteoclasts (OC).

Essentially OC and OB cells are in equilibrium in the absence of MM cells, while
MM and OC cells have a stimulatory effect on each other, and MM cells inhibit OB
and OB cells have little or not effect of MM cells:

Summary of effects of diffusible factors:

     GOC          GOB             GMM
OC  neutral       equilibrium     stimulation
OB  equilibrium   neutral         inhibition
MM  stimulation   neutral         neutral


For this model that has been obtained from Tomlinson (1997), since all the celltypes represent different phenotypes, the fitness functions have been set as follows:
FA = 1 + bz*f_C + ay*f_B            
FB = 1 + ax*f_A - dz*f_C
FC = 1 + bx*f_A

Where:
x = Contribution of OC
y = Contribution of OB
z = Contribution of MM

The WT and all double mutants have been considered as absent


Multiplication factors:

     GOC          GOB             GMM
OC    0            a               b
OB    a            0              -d
MM    b            0               0 

Functions definitions:
```{r}
library(OncoSimulR)

fast_allFitnessEffects <- function(fitnessDF){
  allFitnessEffects(genotFitness = fitnessDF,
                    frequencyDependentFitness = TRUE,
                    frequencyType ='rel')
}

fast_oncoSimulIndiv <- function(fitnessEff){
  oncoSimulIndiv(fitnessEff,
                 model = "McFL",
                 onlyCancer = FALSE,
                 finalTime = 5000,
                 verbosity = 0,
                 mu = 1e-3,
                 initSize = 10000,
                 keepPhylog = TRUE,
                 seed = NULL,
                 detectionProb = NA,
                 detectionSize = NA,
                 errorHitMaxTries = FALSE,
                 errorHitWallTime = FALSE)
}



create_df1 <- function(a, b, c, d, x, y, z){
  data.frame(Genotype = c("WT", "OC", "OB", "A, B", "MM"),
             Fitness = c("0.5", paste("1 +",
               as.character(b*z), " * f_C +", 
               as.character(a*y), " * f_B"),
               paste("1 +",
                 as.character(a*x), " * f_A -",
                 as.character(d*z), " * f_C"),
               "0", 
               paste("1+",
                     as.character(b*x), " * f_A")),
             stringsAsFactors = FALSE)
}
```


The paper describes an initial scenario for which the initial frequencies of the genotypes are not important, for which even with OncoSimulR the frequencies have not been established, the trend for the final frequencies match the paper results:

In multiple myeloma the contribution of the different cells are different. In this scenario, the benefit of diffusible factors that are secreted by MM cells is greater than the benefit that OC cells can obtain through the diffusible factors produced by OB (y < x < z). There exists a polymorphic stable point between MM and OC.

The expected result is that the final frequency of OC will be greater than the final frequency of MM. Although these two seem to be in equilibrium while OB tends to disappear.

```{r}
a <- 1
b <- 2.5
d <- 0.3
x <- 1
y <- 1.2
z <- 1.4

r <- create_df1(a, b, c, d, x, y, z)
afe <- fast_allFitnessEffects(r)
osi_1 <- fast_oncoSimulIndiv(afe)
plot(osi_1, show = "genotypes", type = "line", xlim= c(12, 20))

```

An scenario when all the initial frequencies can also be modeled with this strategy because all the genotypes are being developed with the same ratio of mutation for which it is possible to model that y = x = z.

When a < b an equilibrium OC - OB is created

```{r}
a <- 1
b <- 0.5
d <- 0.3
x <- 1
y <- 1
z <- 1

r <- create_df1(a, b, c, d, x, y, z)
afe <- fast_allFitnessEffects(r)
osi_2 <- fast_oncoSimulIndiv(afe)
plot(osi_2, show = "genotypes", type = "line", xlim= c(12, 40))
```

When b + d > a an equilibrium OC-MM is created

```{r}
a <- 1
b <- 2
d <- 0
x <- 1
y <- 1
z <- 1

r <- create_df1(a, b, c, d, x, y, z)
afe <- fast_allFitnessEffects(r)
osi_3 <- fast_oncoSimulIndiv(afe)
plot(osi_3, show = "genotypes", type = "line", xlim= c(12, 23))
```

In the paper, a third scenario is explained, for which the initial frequencies are important. In this case this model strategy can not be used because the initial frequencies are not being stablished.

